Engineering infinite-range SU($n$) interactions with spin-orbit coupled fermions in an optical lattice

### 2021-07-28 sun_phases/spin_model_paper/main.tex
added minor comment to spin model paper

If the addition $\O(w,\hat X)$ to the SU($n$)-symmetric Hamiltonian $\H_0$ in Eq.~\eqref{eq:H_0} is sufficiently small, namely with operator norm $\norm*{\O(w,\hat X)}$ less than half the spectral gap $\Delta_{\t{gap}}$ of $\H_0$, $\norm*{\O(w,\hat X)}<\Delta_{\t{gap}}/2$, then we can treat the effect of $\O(w,\hat X)$ on the ground-state PS manifold $\E_0$ perturbatively.
The effective Hamiltonians $\H_\eff^{(1)}$ and $\H_\eff^{(2)}$ induced by $\O(w,\hat X)$ on the PS manifold $\E_0$ at leading orders in perturbation theory are \cite{bravyi2011schrieffer}
\begin{align}
  \H_\eff^{(1)} = \P_0 \O(w,\hat X) \P_0,
  &&
  \H_\eff^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \O(w,\hat X) \P_\Delta \O(w,\hat X) \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of $\H_0$ with interaction energy $\Delta$ above that of the PS manifold.
The first order effective Hamiltonian $\H_\eff^{(1)}$ simply projects $\O(w,\hat X)$ onto the PS manifold $\E_0$, and takes the form
\begin{align}
  \H_\eff^{(1)} = \mean w \, \col X,
  \label{eq:H_eff_1}
\end{align}
where the coefficient $\mean w$ is the average of all coefficients $w_k$; and $\col X$ is a collective version of $X$:
\begin{align}
  \mean w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col X \equiv \sum_{k\in\D_N\p{M}} \hat X_k,
  \label{eq:mean_col}
\end{align}
with $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$.
In the case of a magnetic field $\s_\z$ or Ising interactions  $\s_\z\otimes \s_\z$, for example,
\begin{align}
  \col{\s_\z} = \sum_i \s_\z^{(i)} = \S_\z,
  &&
  \col{\s_\z\otimes \s_\z}
  = \sum_{i\ne j} \s_\z^{(i)} \s_\z^{(j)}
  = \S_\z^2 - N \sum_i \s_{\z,i}^2.
\end{align}
The second order effective Hamiltonian $\H_\eff^{(2)}$ in Eq.~\eqref{eq:H_eff} takes more work to simplify due to the presence of a projector $\P_\Delta$ onto the manifold $\E_\Delta$ of states with excitation energy $\Delta$.
This projector essentially picks off the part of $\O(w,\hat X)$ that is strictly off-diagonal with respect to the ground- and excited-state manifolds $\E_0$ and $\E_\Delta$.
We therefore need to decompose $\O(w,\hat X)$ into components that generate states of definite excitation energy when acting on PS states $\ket\psi\in\E_0$.
The SU($n$) symmetry of $\H_0$ enables such a decomposition to take the form
\begin{align}
  \H_0 \O(w,\hat X) \ket\psi
  = \sum_\Delta \p{E_0+\Delta} \O(w^\Delta,\hat X) \ket\psi,
  &&
  E_0 \equiv \sum_{i<j} g_{ij},
  \label{eq:excitation}
\end{align}
where $E_0$ is the interaction energy of PS states, and thinking of the tensor $w$ as a $\abs{\D_N\p{M}}$-component vector, the tensor $w^\Delta$ can be found by
\begin{enumerate*}
\item using the coefficients $g_{ij}$ to construct a matrix $g^{(M)}$ of dimensions $\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
\item projecting $w$ onto the eigenspace of $g^{(M)}$ with eigenvalue $\Delta$.
\end{enumerate*}
We construct $g^{(M)}$ for the single-body ($M=1$) case in Appendix \ref{sec:eigenstates}, and provide explicit forms of $g^{(M)}$ with arbitrary $M$.

@@

Within the permutationally symmetric manifold, we can replace $\s_{\alpha,q}\to\bar s_\alpha\equiv\frac1N\sum_q \s_{\alpha,q}$ at the cost of $O(1/N)$ errors that vanish as $N\to\infty$, so taking this limit we find
\begin{align}
  \ell_\alpha\p{z}
  = \I\p{z} \bar s_\alpha
  + \delta_{\alpha,\tilde\z} \, \xi h,
\end{align}
where
\begin{align}
  \I\p{z} \equiv \lim_{N\to\infty} \f1N \sum_q \f1{z-\sin\p{q}}
  = \f1{2\pi} \int_0^{2\pi} \f{\dd q}{z-\sin\p{q}}
  = \f1{\sqrt{z^2-1}}
  &&
  \t{for}
  &&
  z \notin \sp{-1,1}.
\end{align}
The squared magnitude of the Lax vector is therefore
\begin{align}
  \vec\ell\p{z}^2
  = \sum_\alpha \ell_\alpha\p{z}^2
  = \I\p{z}^2 \sum_{\alpha\ne\tilde\z} \bar s_\alpha^2
  + \sp{\I\p{z} \bar s_{\tilde\z} + \xi h}^2,
\end{align}
where we can define the scalar $Q^2\equiv\sum_\alpha \bar s_\alpha^2$ to simplify
\begin{align}
  \vec\ell\p{z}^2
  = \I\p{z}^2 \p{Q^2 - \bar s_{\tilde\z}^2}
  + \sp{\I\p{z} \bar s_{\tilde\z} + \xi h}^2
  = \I\p{z}^2 Q^2 + \xi^2 h^2
  + 2 \I\p{z} \xi h \bar s_{\tilde\z}.
\end{align}
For initial states with $\bk{\bar s_\z}=0$, we thus find that
\begin{align}
  \vec\ell\p{z}^2 = \f{Q^2}{z^2-1} + \xi^2 h^2,
\end{align}
which is zero when\footnote{Strictly speaking, the zeros in Eq.~\eqref{eq:lax_zeros} occur at values of $z$ at which $\I\p{z}$ is undefined.
We avoid this issue by analytically continuing $\I\p{z}^2$ to the interval $z\in(-1,1)$.}
\begin{align}
  z = \pm \sqrt{1 - \p{\f{Q}{\xi h}}^2}.
  \label{eq:lax_zeros}
\end{align}
These roots change character when $z=0$, suggesting that the critical field $h_\crit$ separating dynamical phases satisfies
\begin{align}
  h_\crit^2 \stackrel{?}{=} \f{Q^2}{\xi^2},
\end{align}
where we use the relation $\stackrel{?}{=}$ to indicate that this ``prediction'' of the Lax analysis is not necessarily valid for all $n$.
For a permutationally symmetric state, up to vanishing $O(1/N)$ corrections we can expand
\begin{align}
  Q^2 = \sum_\alpha \bar s_\alpha^2
  = \sum_{\mu,\nu} \bar s_{\mu\nu} \bar s_{\nu\mu} - \f1n
  = 1 - \f1n
  = \f{n-1}{n},
\end{align}
which implies that
\begin{align}
  h_\crit^2 \stackrel{?}{=} \f{n-1}{n} \times \f{12}{n(n+1)(n-1)}
  = \f{12}{n^2\p{n+1}}.
\end{align}
This Lax analysis correctly predicts that $h_\crit=1$ when $n=2$, but otherwise predicts $h_\crit\sim n^{-3/2}$, which is inconsistent with the finding that $h_\crit\sim n^{-1/3}$ in the mean-field results of the main text (see Figure \ref{fig:crit_fields_X}).

>>>>

If the addition $\O(w,\hat X)$ to the SU($n$)-symmetric Hamiltonian $\H_0$ in Eq.~\eqref{eq:H_0} is sufficiently small, namely with operator norm $\norm*{\O(w,\hat X)}$ less than half the spectral gap $\Delta_{\t{gap}}$ of $\H_0$, $\norm*{\O(w,\hat X)}<\Delta_{\t{gap}}/2$, then we can treat the effect of $\O(w,\hat X)$ on the ground-state PS manifold $\E_0$ perturbatively.
The effective Hamiltonians $\H_\eff^{(1)}$ and $\H_\eff^{(2)}$ induced by $\O(w,\hat X)$ on the PS manifold $\E_0$ at leading orders in perturbation theory are \cite{bravyi2011schrieffer}
\begin{align}
  \H_\eff^{(1)} = \P_0 \O(w,\hat X) \P_0,
  &&
  \H_\eff^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \O(w,\hat X) \P_\Delta \O(w,\hat X) \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of $\H_0$ with interaction energy $\Delta$ above that of the PS manifold.
The first order effective Hamiltonian $\H_\eff^{(1)}$ simply projects $\O(w,\hat X)$ onto the PS manifold $\E_0$, and takes the form
\begin{align}
  \H_\eff^{(1)} = \mean w \, \col X,
  \label{eq:H_eff_1}
\end{align}
where the coefficient $\mean w$ is the average of all coefficients $w_k$; and $\col X$ is a collective version of $X$:
\begin{align}
  \mean w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col X \equiv \sum_{k\in\D_N\p{M}} \hat X_k,
  \label{eq:mean_col}
\end{align}
with $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$.
In the case of a magnetic field $\s_\z$ or Ising interactions  $\s_\z\otimes \s_\z$, for example,
\begin{align}
  \col{\s_\z} = \sum_i \s_\z^{(i)} = \S_\z,
  &&
  \col{\s_\z\otimes \s_\z}
  = \sum_{i\ne j} \s_\z^{(i)} \s_\z^{(j)}
  = \S_\z^2 - N \sum_i \s_{\z,i}^2.
\end{align}
The second order effective Hamiltonian $\H_\eff^{(2)}$ in Eq.~\eqref{eq:H_eff} takes more work to simplify due to the presence of a projector $\P_\Delta$ onto the manifold $\E_\Delta$ of states with excitation energy $\Delta$.
This projector essentially picks off the part of $\O(w,\hat X)$ that is strictly off-diagonal with respect to the ground- and excited-state manifolds $\E_0$ and $\E_\Delta$.
We therefore need to decompose $\O(w,\hat X)$ into components that generate states of definite excitation energy when acting on PS states $\ket\psi\in\E_0$.
The SU($n$) symmetry of $\H_0$ enables such a decomposition to take the form
\begin{align}
  \H_0 \O(w,\hat X) \ket\psi
  = \sum_\Delta \p{E_0+\Delta} \O(w^\Delta,\hat X) \ket\psi,
  &&
  E_0 \equiv \sum_{i<j} g_{ij},
  \label{eq:excitation}
\end{align}
where $E_0$ is the interaction energy of PS states, and thinking of the tensor $w$ as a $\abs{\D_N\p{M}}$-component vector, the tensor $w^\Delta$ can be found by
\begin{enumerate*}
\item using the coefficients $g_{ij}$ to construct a matrix $g^{(M)}$ of dimensions $\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
\item projecting $w$ onto the eigenspace of $g^{(M)}$ with eigenvalue $\Delta$.
\end{enumerate*}
We construct $g^{(M)}$ for the single-body ($M=1$) case below (in Appendix \ref{sec:eigenstates}), and provide explicit forms of $g^{(M)}$ with arbitrary $M$.

@@

Within the permutationally symmetric manifold, we can replace $\s_{\alpha,q}\to\bar s_\alpha\equiv\frac1N\sum_q \s_{\alpha,q}$ at the cost of $O(1/N)$ errors that vanish as $N\to\infty$, so taking this limit we find
\begin{align}
  \ell_\alpha\p{z}
  = \I\p{z} \bar s_\alpha
  + \delta_{\alpha,\tilde\z} \, \xi h,
\end{align}
where
\begin{align}
  \I\p{z} \equiv \lim_{N\to\infty} \f1N \sum_q \f1{z-\sin\p{q}}
  = \f1{2\pi} \int_0^{2\pi} \f{\dd q}{z-\sin\p{q}}
  = \f1{\sqrt{z^2-1}}
  &&
  \t{for}
  &&
  z \notin \sp{-1,1}.
\end{align}
The squared magnitude of the Lax vector is therefore
\begin{align}
  \vec\ell\p{z}^2
  = \sum_\alpha \ell_\alpha\p{z}^2
  = \I\p{z}^2 \sum_{\alpha\ne\tilde\z} \bar s_\alpha^2
  + \sp{\I\p{z} \bar s_{\tilde\z} + \xi h}^2,
\end{align}
where we can define the scalar $Q^2\equiv\sum_\alpha \bar s_\alpha^2$ to simplify
\begin{align}
  \vec\ell\p{z}^2
  = \I\p{z}^2 \p{Q^2 - \bar s_{\tilde\z}^2}
  + \sp{\I\p{z} \bar s_{\tilde\z} + \xi h}^2
  = \I\p{z}^2 Q^2 + \xi^2 h^2
  + 2 \I\p{z} \xi h \bar s_{\tilde\z}.
\end{align}
For initial states with $\bk{\bar s_\z}=0$, we thus find that
\begin{align}
  \vec\ell\p{z}^2 = \f{Q^2}{z^2-1} + \xi^2 h^2,
\end{align}
which is zero when\footnote{Strictly speaking, the zeros in Eq.~\eqref{eq:lax_zeros} occur at values of $z$ at which $\I\p{z}$ is undefined.
We avoid this issue by analytically continuing $\I\p{z}^2$ to the interval $z\in(-1,1)$.}
\begin{align}
  z = \pm \sqrt{1 - \p{\f{Q}{\xi h}}^2}.
  \label{eq:lax_zeros}
\end{align}
These roots change character when $z=0$, suggesting that the critical field $h_\crit$ separating dynamical phases satisfies
\begin{align}
  h_\crit^2 \stackrel{?}{=} \f{Q^2}{\xi^2},
\end{align}
where we use the relation $\stackrel{?}{=}$ to indicate that this ``prediction'' of the Lax analysis is not necessarily valid for all $n$.
For a permutationally symmetric state, up to vanishing $O(1/N)$ corrections we can expand
\begin{align}
  Q^2 = \sum_\alpha \bar s_\alpha^2
  = \sum_{\mu,\nu} \bar s_{\mu\nu} \bar s_{\nu\mu} - \f1n
  = 1 - \f1n
  = \f{n-1}{n},
\end{align}
which implies that
\begin{align}
  h_\crit^2 \stackrel{?}{=} \f{n-1}{n} \times \f{12}{n(n+1)(n-1)}
  = \f{12}{n^2\p{n+1}}.
\end{align}
This Lax analysis correctly predicts that $h_\crit=1$ when $n=2$, but otherwise predicts $h_\crit\sim n^{-3/2}$, which is inconsistent with the finding that $h_\crit\sim n^{-1/3}$ in the mean-field results of the main text (see Figure \ref{fig:crit_fields_X}).
We emphasize that this inconsistency is not a failure of the Lax formalism, but rather an indication that new theoretical tools are necessary to understand multilevel spin models.
